Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-07-22 paper-correlations.tex
some additional changes from kg

\author{Aleksei Samarin\textsuperscript{1,2}}
\author{Vasily Postnicov\textsuperscript{1}}
\author{Marina Karsanina\textsuperscript{1}}
\author{Efim Lavrukhin\textsuperscript{1,2}}
\author{Dina Gafurova\textsuperscript{3}}
\author{Aleksey Khlyupin\textsuperscript{1}}
\author{Kirill Gerke\textsuperscript{1}}
\email{kg@ifz.ru}

\affiliation{\textsuperscript{1}Schmidt Institute of Physics of the Earth of
  Russian Academy of Sciences, Moscow, 107031, Russia}
\affiliation{\textsuperscript{2}Computational Mathematics and Cybernetics,
  Lomonosov Moscow State University, Moscow, 119991, Russia}
\affiliation{\textsuperscript{3}Oil and Gas Research Institute Russian Academy
  of Sciences (OGRI RAS) 3, Gubkina st., Moscow, 119333, Russian Federation}

@@

\section{Background and motivation}
\label{sec:background}
Materials are ubiquitous both in nature and in industrial human activities, and
their physical properties are interconnected to
structure\cite{Torquato_book}\cite{Sahimi_book}\cite{Adler_recon}. The information
about the structure usually comes in the way of 2D and 3D images obtained by
microscopy, tomography and other related methodologies. This structural
information can be used for numerous types of analysis, including simulations of
different processes within studied materials or evaluation of their properties
using so-called pore-scale modelling. Unlike computational modelling techniques
a whole class of approximate methods exists including (rigorous) bounds that
allows very fast estimations of physical properties albeit the bounds themselves
are too wide to be applicable in majority of practical problems. Nonetheless, it
is noteworthy that such bounds are originally based on structural descriptors in
the form of correlation functions.

Two ways exist to obtain correlation functions for a given structure at hand:
measure them either experimentally with the help of scattering intensity or
from digital images. None is perfect, as the first approach suffers from the
limitation of CFs that can be obtained this way, while the second one provides
information with limited resolution or/and resolution to field-of-view ratio. In
addition to the resolution conundrum, the most useful imaging methods such as
X-ray computed tomography (XCT) and scanning electron microscopy (SEM) provide
gray-scale images (e.g., X-ray attenuation or electron back-scattering
distributions) due to their underlying physical principles and require labeling
of constituent phases (this is called segmentation) before computation of
CFs. If properly segmented, high-resolution digital images do provide a
possibility to compute any correlation function and, thus, possibility to
address numerous fundamental and practical research problems.

Correlation functions (CFs) as invaluable universal descriptors of structure are
utilized in a multitude of scientific disciplines: material sciences, rock
physics, soil physics and hydrology, cosmology and food engineering, biology and
many others. Computed from 2D and 3D images, CFs can be applied to:
\begin{enumerate}
  \item to characterize the morphology and representativeness via correlation
    lengths;
  \item to perform stochastic reconstructions from experimentally measured CFs
    or 2D to 3D reconstructions based on CFs that can be extracted from less
    dimensions or through penetrable sphere model;
  \item compare different structures to each other, including verification of
    stochastic reconstructions;
  \item compress structural information in the form of raw or parameterized CFs
    with the possibility to recover the structure using stochastic
    reconstruction
  \item describe structural dynamics under different boundary conditions;
  \item extract structural features for deep learning;
  \item fuse multi-scale images and structural information as obtained by
    different methods into a single digital model;
  \item quantify spatial heterogeneity.
\end{enumerate}

Stochastic reconstruction is a separate huge topic, as this approach allows to
solve an inverse problem and recover structure from known set of correlation
functions. This ability for recovery serves as the basis for majority of usages
in the list above. While it is necessary to compute CFs from digital images as a
target set for stochastic reconstructions, they rely more on efficient
recomputations or optimizations, e.g., re-evaluation during simulated
annealing. Such optimizations are not the part of the computational package and
are not considered in this work.

There are different types of correlation functions that in general describe some
probabilities. The main parameter of any correlation function is the number of
points that are utilized to evaluate such a probability. While so-called
$n$-point probability function\cite{Torquato_book} will totally describe any
structure in the limit of $n \rightarrow$ number of voxels/pixels on the image,
computing or storing such a function is not practical. Based on information
content, it was shown that increasing the number of point $n > 2$ only
marginally improves the quality of the structure quantification and these
improvements decay with increasing $n$. For stochastic reconstruction purposes
computation of higher-order statistics will, arguably, not be balanced by
increased number of points, but this is the topic of active research. For
aforementioned reasons, we mainly focus on 2-point statistics, leaving the
possibility to include higher order CFs in the future work. The simplest 2-point
probability function ($S_2$ or autocorrelation) actually arise from small angle
scattering experiments and measures the probability that both ends of line
segment with a given length fall into the same phase. Other types include the
probability of a whole line segment to fall into the phase ($L_2$ function) or
to lie on the interface between phases ($F_{ss}$ function), and mainly originate
from aforementioned bounds on physical properties such as permeability and
elasticity. The general idea of computation of different 2-point CFs is shown in
\cref{fig:functions}. All major details and numerous analytical cases can be
found in seminal work of Torquato \cite{Torquato_book}. To all CFs described in this
book and applicable to digital images we shall refer as to classical correlation
functions.

@@

\section{Introduction to the correlation functions}
\label{sec:math}
In practice porous media are often described with special descriptors called
``correlation functions''. In this section we provide a brief introduction to
these functions which can also be found in \cite{Torquato_book}. The functions
supported by our package are:
\begin{itemize}
\item Lineal-path function $L_2$.
\item Two-point function $S_2$.
\item Cluster function $C_2$.
\item Surface-surface function $F_{ss}$.
\item Surface-void function $F_{sv}$.
\item Pore size function $P$.
\item Chord length function $p$.
\item Phase cross-correlation function $\rho_{ij}$
\end{itemize}

One of basic correlation functions is a lineal path function which is defined as
follows for homogenous media:
\begin{definition}
  Lineal path function $L_2^{(i)}(\bm{r})$ equals to probability that a
  vector $\bm{r}$ lies wholly in phase $i$ when randomly thrown into the
  sample.
\end{definition}
If a sample is isotropic a vector $\bm{r}$ can be replaced with its length
$r$ and the definition becomes:
\begin{definition}
  Lineal path function (for isotropic media) $L_2^{(i)}(r)$ equals to
  probability that a line segment of length $r$ lies wholly in phase $i$ when
  randomly thrown into the sample.
\end{definition}

Another function of big importance is a two-point correlation function $S_2$. A
definition for homogenous media is as follows:
\begin{definition}
  Two point function $S_2^{(i)}(\bm{r})$ equals to probability that both ends
  of a vector $\bm{r}$ lie in phase $i$ when the vector is randomly thrown
  into the sample.
\end{definition}
For isotopic media a vector $\bm{r}$ can be again reduced to a scalar value
$r$ (i.e. a correlation function depends only on length of a vector and not on
its direction.) Mathematically this definition can be expressed with the
following equation:
\begin{equation}
  S_2^{(i)}(\bm{r}) = \langle I^{(i)}(\bm{x}) I^{(i)}(\bm{x} +
  \bm{r}) \rangle
  \label{eq:s2-def}
\end{equation}
where $I^{(i)}$ is an indicator function for a set of all points belonging to
the phase $i$ and $\langle \cdots \rangle$ is an average over all xs.

A function closely related to $S_2$ is phase cross-correlation function
$\rho_{ij}$ which is defined with the following equation:
\begin{equation}
  \rho_{ij}(\bm{r}) = \langle I^{(i)}(\bm{x}) I^{(j)}(\bm{x} +
  \bm{r}) \rangle
  \label{eq:cross-def}
\end{equation}

The next function is a cluster function $C_2$.
\begin{definition}
  Cluster function $C_2^{(i)}(\bm{r})$ equals to probability that both
  ends of a vector $\bm{r}$ lie in the same cluster when the vector is
  randomly thrown into the sample. A cluster is a set of points of the same
  phase $i$ where any two points can be connected by a path lying entirely in
  that phase.
\end{definition}

The next two functions are called surface-surface ($F_{ss}$) and surface-void
($F_{sv}$) functions and describe surface of a sample. We skip strict verbal
definitions and introduce an interface indicator function instead:
\begin{equation*}
  M^{(i)}(\bm{r}) = | \nabla I^{(i)}(\bm{r}) |
\end{equation*}
Here differentiation is understood in the sense of generalized functions,
because an ordinary differential of $I$ is either zero or undefined. For
homogenous media definitions of these functions are:
\begin{align}
  F_{ss}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) M^{(i)}(\bm{x} +
  \bm{r}) \rangle \label{eq:fss-def} \\
  F_{sv}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) I^{(void)}(\bm{x}
  + \bm{r}) \rangle \label{eq:fsv-def}
\end{align}
These functions are defined for media with dimensionality of 2 and
higher. $F_{sv}$ is a physical quantity inversely proportional to length and
$F_{ss}$ is inversionaly proportional to surface area, therefore we can use
$\mu m^{-1}$ to measure surface-void and $\mu m^{-2}$ to measure surface-surface
functions. Putting it simply, surface-void function measures length (2D case) or
surface (3D case) of a part of the interface which lies in the void phase of
porous medium shifted by vector $\bm{r}$. Surface-surface function characterizes
intersections of interfaces of porous medium and its shifted version. As always,
there are versions of these functions applicable to scalar argument for
isotropic media.

@@

Finally, a chord length function $p^{(i)}(r)$ is defined as follows:
\begin{definition}
$p^{(i)}(r)dz$ = Probability of finding a chord of length between $r$ and $r+dr$
in phase $i$. Chord is a line segment which lies entirely in the same phase and
touches the interface with its ends.
\end{definition}

\verb+CorrelationFunctions.jl+ is capable to compute all these
functions. Usually there are two implementations for a function. The first
implementation computes a correlation function in specific predefined directions
(i.e. orientation of test vectors thrown into a sample is fixed). There are one
predefined direction in 1D case, four directions in 2D case and thirteen
directions in 3D case. This implementation is somewhat slow but requires much
less memory than the second one. The second implementation computes a
correlation function in all possible directions and is faster but requires much
memory. These implementations are contained in \code{Directional} and \code{Map}
modules of the package correspondingly.

These functions can work in periodic and non-periodic modes. Periodic mode
assumes that an input sample is periodically continued to the whole Euclidean
space. Non-periodic mode assumes zero-padded input.

\section{The map approach.}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of interface indicator $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into the sample). If we want to calculate $S_2$ function in periodic mode
we need only to apply \cref{eq:s2-ft} to an input to obtain the result. In
non-periodic mode we need to pad an input with zeros to at least twice the size
minus one for each dimension. In periodic mode a total number of trials is
independent of $r$ and equals to the number of elements in an input array. In
non-periodic mode $N(r)$ is calculated as described in
\cref{sec:number-of-trials}. The following algorithm summarizes all said above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
  \State $A \gets zeropad(A)$
  \Comment Pad with at least $2\cdot size(A, i) - 1$ zeros for each dimension $i$.
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}| / N$
  \Comment $N$ is an array with numbers of trials
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation if $O(M \log M)$ where $M$ is the
number of elements in the input.

This function is a special case of a cross-correlation function $\rho_{ij}$
which computes a correlation between phases $i$ and $j$ using a formula
\begin{equation}
  \rho_{ij}(r) = \frac{F^{-1} [F[A^{(i)}](z) \overline{F[A^{(j)}](z)}] (r)}{N(r)} \label{eq:cross-ft}
\end{equation}
or an equivalent algorithm
\begin{algorithmic}[1]
  \Procedure{$\rho_{ij}$}{$A, i, j, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(i)} \gets I^{(i)} (A)$
  \State $A^{(j)} \gets I^{(j)} (A)$
  \State $\hat{A}^{(i)} \gets F[A^{(i)}]$
  \State $\hat{A}^{(j)} \gets F[A^{(j)}]$
  \State $\hat{\rho} \gets \hat{A}^{(i)} \overline{\hat{A}^{(j)}} / N$
  \State \textbf{return} $F^{-1} [\hat{\rho}]$
  \EndProcedure
\end{algorithmic}

Cluster function can be calculated in the similar manner. Firstly, we apply an
indicator function to an input $A$: $A^{(i)} \xleftarrow{I^{(i)}} A$. When we
label connected components in $A^{(i)}$ using one of many popular algorithms for
this purpose \textcolor{red}{cite something}. This algorithm must take into
account if the function is calculated in periodic mode or not. Then $C_2$
function is a sum of $S_2$ functions calculated for all clusters
separately. Algorithmic complexity of this implementation is $O(MC \log M)$
where $M$ is the number of elements in the input and $C$ is the number of
clusters. The algorithm is as follows:
\begin{algorithmic}[1]
  \Procedure{$C_2$}{$A, phase, periodic$}
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $CC \gets label\_components(A^{(phase)}, periodic)$
  \State $N \gets maximum(CC)$ \Comment Number of connected components
  \State \textbf{return} $\sum\limits_{n=1}^N S_2(CC, n, periodic)$
  \EndProcedure
\end{algorithmic}

For surface-surface and surface-void functions we use an edge detection filter
and then calculate either autocorrelation of the edge or cross-correlation of
the edge and the void phase. The edge is extracted by convolving an input with a
short high-pass filter $H$. It has a width $7$ and all its coefficients with
exception of the central coefficient are inversely proportional to distance
from the center:
\begin{equation}
  \begin{aligned}
    H_{ij} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2}} & \quad i \ne 3, j \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right. \\
    H_{ijk} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2 + (k-3)^3}} & \quad i \ne 3, j \ne 3, k
      \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right.
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $C$ is such that all coefficients of $H'$ sum to zero, and $S$ equals to
$30.45849$ in 2D and $172.96232$ in 3D case.

@@

Our library does not have an implementation for lineal-path maps, although we
know that such implementations exist.

All these algorithms contain parallelizable operations such as fast Fourier
transform, connected components labeling and image filtering, hence they are
perfectly suited for execution on GPUs.

\section{The scan approach.}
\label{sec:scan}
All algorithms described above are executed on the whole input and hence require
a lot of memory. The scan approach deals with memory limitations at the cost of
higher execution time. In this approach we select a list of predefined
directions and cut all one-dimensional slices from the input along those
directions. Then for each direction we use an appropriate algorithm in
\cref{sec:map} to calculate a correlation function just for that slice. When all
slices are processed we calculate the result using the formula:
\begin{equation*}
  F^{\bm{d}}(r) = \frac{\sum\limits_i F^{\bm{d}}_i(r) N_i(r)}{\sum\limits_i N_i(r)}
\end{equation*}
Here $F^{\bm{d}}(r)$ is a correlation function computed with respect to a
direction $\bm{d}$, $F^{\bm{d}}_i(r)$ is a ``partial'' correlation
function calculated for a $i$-th slice and $N_i(r)$ is a total number of trials
for a correlation length $r$ and slice $i$. Some algorithms require
preprocessing stages before cutting the input in slices, e.g. image filtering
for edge detection or connected components labeling is performed on the whole
array, not on slices.

>>>>

\author{Vasily Postnicov\textsuperscript{1}}
\author{Aleksei Samarin\textsuperscript{1,2}}
\author{Marina Karsanina\textsuperscript{1}}
\author{Aleksey Khlyupin\textsuperscript{1}}
\author{Kirill Gerke\textsuperscript{1}}
\email{kg@ifz.ru}

\affiliation{\textsuperscript{1}Schmidt Institute of Physics of the Earth of
  Russian Academy of Sciences, Moscow, 107031, Russia}
\affiliation{\textsuperscript{2}Computational Mathematics and Cybernetics,
  Lomonosov Moscow State University, Moscow, 119991, Russia}

@@

\section{Background and motivation}
\label{sec:background}
Materials are ubiquitous both in nature and in industrial human activities, and
their physical properties are interconnected to
structure \cite{Torquato_book,Sahimi_book}. The information
about the structure usually comes in the way of 2D and 3D images obtained by
microscopy \cite{moussaoui2018,neumann2019,FIB-SEMpaper}, tomography \cite{xctmat_book} and other related methodologies.
This structural information can be used for numerous types of analysis, including simulations of
different processes within studied materials or evaluation of their properties
based on 3D structural data \cite{youssef2005,Miao2017,FDMSS}. Unlike computational modelling techniques
a whole class of approximate methods exists including (rigorous) bounds that
allows very fast estimations of physical properties \cite{eshelby1957,berryman1986use,rozanski2023} albeit the bounds themselves
are too wide to be applicable in the majority of practical problems. Nonetheless, it
is noteworthy that many of such bounds are originally based on structural descriptors in
the form of correlation functions (CFs).

Two ways exist to obtain correlation functions for a given structure at hand:
measure them either experimentally with the help of scattering intensity \cite{debye1957scattering,li2018direct} or
from digital images \cite{berryman1985measurement,ma2018SS}. None is perfect, as the first approach suffers from the
limitation of CFs that can be obtained this way, while the second one provides
information with limited resolution or/and resolution to field-of-view ratio \cite{gerke2015universal}. In
addition to the resolution conundrum, the most useful imaging methods such as
X-ray computed tomography (XCT) and scanning electron microscopy (SEM) provide
gray-scale images (e.g., X-ray attenuation or electron back-scattering
distributions) due to their underlying physical principles and require labeling
of constituent phases (or segmentation) before computation of
CFs \cite{samarin2023robust}. If properly segmented \cite{NNseg}, high-resolution digital images do provide a
possibility to compute any correlation function and, thus, possibility to
address numerous fundamental and practical research problems.

Correlation functions as invaluable universal descriptors of structure are
utilized in a multitude of scientific disciplines: material sciences \cite{hasanabadi20163d,Havelka,feng2018reconstruction,chen2022quantifying}, rock
physics \cite{ledesma2018effect}, soil physics and hydrology \cite{PLoS_ONE,KarsaninaEJSS}, cosmology and food engineering \cite{TakadaJain,Derossi2019}, biology \cite{veatch2012correlation} and
many others. Computed from 2D and 3D images, CFs can be applied to:
\begin{enumerate}
  \item to characterize the morphology and representativeness via correlation
    lengths \cite{vcapek2009stochastic,thovert2011grain,tensorPRE};
  \item to perform stochastic reconstructions from experimentally measured CFs
    or 2D to 3D reconstructions based on CFs that can be extracted from less
    dimensions or through penetrable sphere model \cite{Adler_recon,Y-T,tahmasebiPRL,hasanabadi20163d,karsaninaPRL};
  \item compare different structures to each other, including verification of
    stochastic reconstructions \cite{vcapek2009stochastic,tahmasebiPRL,EPL2};
  \item compress structural information in the form of raw or parameterized CFs
    with the possibility to recover the structure using stochastic
    reconstruction \cite{SciRep1,Havelka,KarsaninaEJSS};
  \item describe structural dynamics under different boundary conditions \cite{chen2015dynamic,xu2022correlation,fomin2023soil};
  \item extract structural features for deep learning \cite{pilania2017multi,kamrava2020linking,roding2020predicting,KarsaninaEJSS};
  \item fuse multi-scale images and structural information as obtained by
    different methods into a single digital model \cite{Geoderma2018};
  \item quantify spatial heterogeneity \cite{zhang2000pore,REVpaper}.
\end{enumerate}

Stochastic reconstruction is a separate huge topic, as this approach allows to
solve an inverse problem and recover structure from known set of correlation
functions \cite{Adler_recon,Y-T,vcapek2009stochastic,hasanabadi20163d,Havelka,feng2018reconstruction,tahmasebiPRL,EPL2,thovert2011grain,tensorPRE}. This ability for recovery serves as the basis for majority of usages
in the list above. While it is necessary to compute CFs from digital images as a
target set for stochastic reconstructions, they rely more on efficient
recomputations or optimizations, e.g., re-evaluation during simulated
annealing. Such optimizations are not the part of the computational package described here and
are not considered in this work.

There are different types of correlation functions that in general describe some
probabilities. The main parameter of any correlation function is the number of
points that are utilized to evaluate such a probability. While so-called
$n$-point probability function \cite{Torquato_book} will completely describe any
structure in the limit of $n \rightarrow$ number of voxels/pixels on the image,
computing or storing such a function is not practical. Based on information
content, it was shown that increasing the number of point $n > 2$ only
marginally improves the quality of the structure quantification and these
improvements decay with increasing $n$ \cite{yao1993high,Gommes2}. For stochastic reconstruction purposes
computation of higher-order statistics will, arguably, not be balanced by
increased number of points, but this is the topic of active research. For
aforementioned reasons, we mainly focus on 2-point statistics, leaving the
possibility to include higher order CFs in the future work. The simplest 2-point
probability function ($S_2$ or autocorrelation) actually arise from small angle
scattering experiments and measures the probability that both ends of line
segment with a given length fall into the same phase. Other types include the
probability of a whole line segment to fall into the phase ($L_2$ function) or
to lie on the interface between phases ($F_{ss}$ function), and mainly originated
from aforementioned bounds on physical properties such as permeability and
elasticity. The general idea of computation of different 2-point CFs is shown in
\cref{fig:functions}. All major details and numerous analytical cases can be
found in seminal work of Torquato \cite{Torquato_book}. To all CFs described in this
book and applicable to digital images we shall refer to as classical correlation
functions.

@@

\section{A brief introduction to correlation functions}
\label{sec:math}
In this section we shall provide major properties of correlation functions
supported by our package; the full list of functions is 
(see \cref{fig:functions} for graphical explanation in addition to definitions below):
\begin{itemize}
\item Two-point probability function $S_2$,
\item Lineal-path function $L_2$,
\item Cluster function $C_2$,
\item Surface-surface function $F_{ss}$,
\item Surface-void function $F_{sv}$,
\item Pore size function $P$,
\item Chord length function $p$,
\item Phase cross-correlation function $\rho_{ij}$.
\end{itemize}
Note that cross-correlation $\rho_{ij}$ is actually not a classical CF, but technically is a variety of $S_2$
for the case when the end of line segment lie in different phases. This function is useful for multi-phase
structures only, as for binary media two-point probability CFs are interdependent.

The most widely used is a two-point probability correlation function $S_2$, or autocorrelation. A
definition for homogenous media is as follows:
\begin{definition}
  Two point function $S_2^{(i)}(\bm{r})$ equals to probability that both ends
  of a vector $\bm{r}$ lie in phase $i$ when the vector is randomly thrown
  into the sample.
\end{definition}
For isotopic media a vector $\bm{r}$ can be reduced to a scalar value
$r$ (i.e., a correlation function depends only on length of a vector and not on
its direction). Mathematically this definition can be expressed with the
following equation:
\begin{equation}
  S_2^{(i)}(\bm{r}) = \langle I^{(i)}(\bm{x}) I^{(i)}(\bm{x} +
  \bm{r}) \rangle
  \label{eq:s2-def}
\end{equation}
where $I^{(i)}$ is an indicator function for a set of all points belonging to
the phase $i$ and $\langle \cdots \rangle$ is ensemble average.

A function closely related to $S_2$ is phase cross-correlation function
$\rho_{ij}$ which is defined as follows:
\begin{equation}
  \rho_{ij}(\bm{r}) = \langle I^{(i)}(\bm{x}) I^{(j)}(\bm{x} +
  \bm{r}) \rangle
  \label{eq:cross-def}
\end{equation}

Another basic correlation function is a lineal path function. It provides some non-trivial
information (albeit usually not complete \cite{vcapek2011transport}) about the connectedness
of the phase and is defined for homogenous media as:
\begin{definition}
  Lineal path function $L_2^{(i)}(\bm{r})$ equals to probability that a
  vector $\bm{r}$ lies wholly in phase $i$ when randomly thrown into the
  sample.
\end{definition}
If a sample is isotropic a vector $\bm{r}$ can be replaced with its length
$r$ and the definition becomes:
\begin{definition}
  Lineal path function (for isotropic media) $L_2^{(i)}(r)$ equals to
  probability that a line segment of length $r$ lies wholly in phase $i$ when
  randomly thrown into the sample.
\end{definition}

The function that contain a lot of structural connectivity information \cite{JiaoPNAS}
is cluster function $C_2$.
\begin{definition}
  Cluster function $C_2^{(i)}(\bm{r})$ equals to probability that both
  ends of a vector $\bm{r}$ lie in the same cluster when the vector is
  randomly thrown into the sample. A cluster is a set of points of the same
  phase $i$ where any two points can be connected by a path lying entirely in
  that phase.
\end{definition}
Note that for a fully connected phase $C_2$ will be equal to $S_2$, something
we shall utilize later for computations.

The next two functions are called surface-surface ($F_{ss}$) and surface-void
($F_{sv}$) functions and describe an interphase between phases. We skip strict verbal
definitions and first introduce an interface indicator function instead:
\begin{equation*}
  M^{(i)}(\bm{r}) = | \nabla I^{(i)}(\bm{r}) |
\end{equation*}
Here differentiation is understood in the sense of generalized functions,
because an ordinary differential of $I$ is either zero or undefined. For
homogenous media the definitions of surface functions then are:
\begin{align}
  F_{ss}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) M^{(i)}(\bm{x} +
  \bm{r}) \rangle \label{eq:fss-def} \\
  F_{sv}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) I^{(void)}(\bm{x}
  + \bm{r}) \rangle \label{eq:fsv-def}
\end{align}
These functions are defined for media with dimensionality of 2 and
higher. $F_{sv}$ is a physical quantity inversely proportional to length and
$F_{ss}$ is inversionaly proportional to surface area. Putting it simply, 
surface-void function measures length (2D case) or surface (3D case) of a part of the
interface which lies in the void phase of porous medium shifted by vector $\bm{r}$. 
Surface-surface function characterizes intersections of interfaces of porous medium and its shifted version.
As always, there are versions of these functions applicable to scalar argument for
isotropic media.

@@

Finally, a chord length function $p^{(i)}(r)$ is defined as:
\begin{definition}
$p^{(i)}(r)dz$ = Probability of finding a chord of length between $r$ and $r+dr$
in phase $i$. Chord is a line segment which lies entirely in the same phase and
touches the interface with its ends.
\end{definition}

More detailed information on all classical CFs can be found in \cite{Torquato_book}.

\section{Computational methodology}

\verb+CorrelationFunctions.jl+ is capable to compute all aforementioned correlation
functions. Where possible we provide two implementations for each CF. The first
implementation computes a correlation function in specific predefined directions
(i.e., orientation of test vectors thrown into a sample is fixed) \cite{jiao2014chawla,EPL1}.
There are one predefined direction in 1D case, four directions in 2D case 
(two orthogonal and two diagonal) and thirteen directions in 3D case (three orthogonal, 
six diagonal within each orthogonal plane and four diagonal). The line segment scanning
approach is slower for some CFs but requires much less memory resources. The full map approach
establishes correlations between all points on the image is usually faster. 
Both implementations are contained in \code{Directional} and \code{Map}
modules of the package, correspondingly.

With the help of the package one can compute CFs in periodic and non-periodic modes. 
Periodic mode assumes that an input sample is periodically continued to the whole Euclidean
space. Non-periodic mode assumes zero-padded input.

\subsection{Full correlation map}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of interface indicator $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into the sample). If we want to calculate $S_2$ function in periodic mode
we need only to apply \cref{eq:s2-ft} to an input to obtain the result. In
non-periodic mode we need to pad an input with zeros to at least twice the size
minus one for each dimension. In periodic mode a total number of trials is
independent of $r$ and equals to the number of elements in an input array. In
non-periodic mode $N(r)$ is calculated as described in
\cref{sec:number-of-trials}. The following algorithm summarizes all the above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
  \State $A \gets zeropad(A)$
  \Comment Pad with at least $2\cdot size(A, i) - 1$ zeros for each dimension $i$.
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}| / N$
  \Comment $N$ is an array with numbers of trials
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation is $O(M \log M)$, where $M$ is the
number of elements in the input image.

Computation of $S_2$ is a special case of a cross-correlation function $\rho_{ij}$
which computes a correlation between phases $i$ and $j$ using a formula:
\begin{equation}
  \rho_{ij}(r) = \frac{F^{-1} [F[A^{(i)}](z) \overline{F[A^{(j)}](z)}] (r)}{N(r)} \label{eq:cross-ft}
\end{equation}
that algorithmically operates as follows:
\begin{algorithmic}[1]
  \Procedure{$\rho_{ij}$}{$A, i, j, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(i)} \gets I^{(i)} (A)$
  \State $A^{(j)} \gets I^{(j)} (A)$
  \State $\hat{A}^{(i)} \gets F[A^{(i)}]$
  \State $\hat{A}^{(j)} \gets F[A^{(j)}]$
  \State $\hat{\rho} \gets \hat{A}^{(i)} \overline{\hat{A}^{(j)}} / N$
  \State \textbf{return} $F^{-1} [\hat{\rho}]$
  \EndProcedure
\end{algorithmic}

Cluster function can be calculated with the help of previuously described procedures. 
Firstly, we apply an indicator function to an input $A$: $A^{(i)} \xleftarrow{I^{(i)}} A$. When we
label connected components in $A^{(i)}$ using a function that is a part of Julia \code{ImageMorphology.jl} library. 
Labeling algorithm must take into account if the function is calculated with periodic boundary 
conditions \cite{evstigneev2023}. For periodic boundary conditions we implemented our own solution 
as part of the package, as it is absent in the public domain. Then $C_2$
function is a sum of $S_2$ functions calculated for all clusters
separately. Algorithmic complexity of this implementation is $O(MC \log M)$
where $M$ is the number of elements in the input and $C$ is the number of
clusters. The algorithm is then written as:
\begin{algorithmic}[1]
  \Procedure{$C_2$}{$A, phase, periodic$}
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $CC \gets label\_components(A^{(phase)}, periodic)$
  \State $N \gets maximum(CC)$ \Comment Number of connected components
  \State \textbf{return} $\sum\limits_{n=1}^N S_2(CC, n, periodic)$
  \EndProcedure
\end{algorithmic}

For surface-surface and surface-void functions we use an edge detection filter \cite{samarin2023robust}
and then calculate either autocorrelation of the edge or cross-correlation of
the edge and the void phase. The edge is extracted by convolving an input with a
short high-pass filter $H$. It has a width $7$ and all its coefficients with
exception of the central coefficient are inversely proportional to distance
from the center:
\begin{equation}
  \begin{aligned}
    H_{ij} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2}} & \quad i \ne 3, j \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right. \\
    H_{ijk} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2 + (k-3)^3}} & \quad i \ne 3, j \ne 3, k
      \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right.
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $C$ is such that all coefficients of $H'$ sum to zero, and $S$ equals to
$30.45849$ in 2D and $172.96232$ in 3D case. The parameters of the filter were adjusted
based on comparison against exact computations of surface functions on smooth-boundary
sets \cite{Postnicov2023}.

@@

Our library does not have an implementation for lineal-path maps, although we
know that such implementations exist \cite{Paper}. The reason is a very slow
computation even on GPUs, plus the resulting map is not for all points on the image,
but within a spherical volume (due to rotation of the 3D image during lineal-path
evaluation). Moreover, linear scan approach is orders of magnitude faster for $L_2$.

All described algorithms contain parallelizable operations such as fast Fourier
transform, connected components labeling and image filtering, hence they are
perfectly suited for execution on GPUs and provide very fast computations (see wall times below).

\subsection{Directional scanning with line segment}
\label{sec:scan}
All algorithms described above (full correlation maps) are executed on the whole input and hence require
a lot of RAM memory. The scan approach deals with memory limitations at the cost of
higher execution time (except for $L_2$). With line-segment scanning we select a list of predefined
directions and cut all one-dimensional slices from the input along those
directions. Then for each direction we use an appropriate algorithm in
\cref{sec:map} to calculate a correlation function just for that single slice. When all
slices are processed we calculate the resulting directional or averaged CF using the formula:
\begin{equation*}
  F^{\bm{d}}(r) = \frac{\sum\limits_i F^{\bm{d}}_i(r) N_i(r)}{\sum\limits_i N_i(r)}
\end{equation*}
Here $F^{\bm{d}}(r)$ is a correlation function computed with respect to a
direction $\bm{d}$, $F^{\bm{d}}_i(r)$ is a ``partial'' correlation
function calculated for a $i$-th slice and $N_i(r)$ is a total number of trials
for a correlation length $r$ and slice $i$. Some algorithms require
preprocessing stages before cutting the input in slices, e.g., image filtering
for edge detection or connected components labeling performed on the whole
array, not on slices.
